---
title: "Understanding the Opioid Epidemic Through Data Science: A Regional and Demographic Analysis"
author: Isabella Villanueva
format:
   html:
     embed-resources: true
---
### Description
Welcome to my data science portfolio, where I apply statistical and analytical skills to uncover critical insights into one of the most pressing public health crises in recent U.S. history. Using a comprehensive dataset from the Centers for Disease Control and Prevention (CDC) on drug overdose death rates (1999–2016), this project explores key questions about how drug mortality has evolved over time, the demographic factors at play, and regional disparities in the epidemic.  

Through interactive visualizations and in-depth analysis, you’ll discover:  

1.  Which regions of the U.S. have seen the sharpest increases in drug poisoning death rates over time.  
2.  Whether Hispanic origin is strongly associated with drug mortality across the nation.  
3.  How drug poisoning death rates differ between men and women in various U.S. regions.  
4.  Which age ranges see the most drug poisoning deaths?

This project not only demonstrates the power of data in addressing real-world challenges but also serves as a foundation for my growing expertise in public health analytics. Dive deeper into the insights on the full project page!  

### Link to Download Written Report
[View the Full Analysis in My Written Report](https://github.com/bellavillan/PM-566-Final-Project/blob/main/Written-Report.pdf)

## Interactive Data Visualizations: Evolving Trends of the Opioid Crisis
These four compelling interactive plots will unravel the complexities of drug poisoning deaths across the United States from 1999 to 2016. These visualizations bring data to life, answering critical questions about regional trends, Hispanic origin and mortality, gender disparities, and age vulnerabilities. 

Through clear and engaging graphs, these interactive plots uncover hidden insights, challenge assumptions, and guide viewers toward a deeper understanding of one of the most pressing public health crises of our time.

```{r setup and data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
drug_mortality <- read_csv("~/Downloads/NCHS_-_Drug_Poisoning_Mortality_by_State__United_States_20241018.csv", show_col_types = FALSE)
```

```{r states categories, include=FALSE}
library(dplyr)

# Define the regions into categories
northeast <- c('Connecticut', 'Maine', 'Massachusetts', 'New Hampshire', 'New Jersey', 'New York', 'Pennsylvania', 'Rhode Island', 'Vermont')
south <- c('Alabama', 'Arkansas', 'Delaware', 'Florida', 'Georgia', 'Kentucky', 'Louisiana', 'Maryland', 'Mississippi', 'North Carolina', 'Oklahoma', 'South Carolina', 'Tennessee', 'Texas', 'Virginia', 'West Virginia', 'District of Columbia')  # Added D.C.to South region
midwest <- c('Illinois', 'Indiana', 'Iowa', 'Kansas', 'Michigan', 'Minnesota', 'Missouri', 'Nebraska', 'North Dakota', 'Ohio', 'South Dakota', 'Wisconsin')
west <- c('Alaska', 'Arizona', 'California', 'Colorado', 'Hawaii', 'Idaho', 'Montana', 'Nevada', 'New Mexico', 'Oregon', 'Utah', 'Washington', 'Wyoming')

# Categorize states into regions
drug_mortality <- drug_mortality %>%
  mutate(Region = case_when(
    State %in% northeast ~ "Northeast",
    State %in% south ~ "South",
    State %in% midwest ~ "Midwest",
    State %in% west ~ "West",
    State == "United States" ~ "National",  # Handle "United States" separately
    TRUE ~ "Unknown"
  ))

# Convert Region to a factor (categorical variable)
drug_mortality$Region <- factor(drug_mortality$Region, levels = c("Northeast", "South", "Midwest", "West", "National"))
```
```{r faceted plot, echo=FALSE}
# Calculate national mean death rate by year using 'US Crude Rate'
national_death_rate <- drug_mortality %>%
  group_by(Year) %>%
  summarize(mean_death_rate = mean(`US Crude Rate`, na.rm = TRUE), .groups = 'drop')

# Group by Year and Region, calculate the mean death rate by region
mean_death_rate_by_region <- drug_mortality %>%
  filter(Region != "National") %>%
  group_by(Year, Region) %>%
  summarize(mean_death_rate = mean(`Crude Death Rate`, na.rm = TRUE), .groups = 'drop')

# Combine national and regional data
mean_death_rate_combined <- bind_rows(
  mean_death_rate_by_region %>% mutate(Source = "Region"),
  national_death_rate %>% mutate(Source = "National", Region = "National")  # Label National data as a region
)
```

```{r plot 1- all, echo=FALSE}
# Install and load necessary libraries
suppressWarnings(suppressMessages({
  if (!require(plotly)) install.packages("plotly")
  library(plotly)
  if (!require(dplyr)) install.packages("dplyr")
  library(dplyr)
}))

# Assign colors to each region and red for National
region_colors <- c("Northeast" = "darkorange2", "South" = "#8494FF", 
                   "Midwest" = "aquamarine3", "West" = "#ED68ED", "National"= "red")

# Create separate datasets for National and Region
national_data <- subset(mean_death_rate_combined, Source == "National")
region_data <- subset(mean_death_rate_combined, Source == "Region")

# Create the interactive plot
mean_death_rate_region_nation <- plot_ly()

# Add the National line
mean_death_rate_region_nation <- mean_death_rate_region_nation %>%
  add_lines(
    data = national_data,
    x = ~Year,
    y = ~mean_death_rate,
    name = "National",  # Legend entry
    line = list(color = "red", width = 3),  # Red line
    hoverinfo = "text",
    text = ~paste("Year:", Year, "<br>Mean Rate:", mean_death_rate, "<br>Region: National")
  )

# Add the regional lines
mean_death_rate_region_nation <- mean_death_rate_region_nation %>%
  add_lines(
    data = region_data,
    x = ~Year,
    y = ~mean_death_rate,
    color = ~Region,  # Map colors to Region
    colors = region_colors,
    name = ~Region,  # Legend entries for regions
    line = list(color = region_colors, width = 2),  
    hoverinfo = "text",
    text = ~paste("Year:", Year, "<br>Mean Rate:", mean_death_rate, "<br>Region:", Region),
    opacity = 0.7  # Transparency for regional lines
  )

# Customize the layout
mean_death_rate_region_nation <- mean_death_rate_region_nation %>%
  layout(
    title = "Fig. 1: Mean Drug Poisoning Death Rate by Region and National (1999-2016)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Mean Death Rate (Deaths per 100,000)"),
    legend = list(title = list(text = "Region"))
  )

# Display the plot
mean_death_rate_region_nation
```
**Figure 1 Description:** This graph shows drug poisoning death rates (per 100,000) from 1999 to 2016 across U.S. regions compared to the national average (red line). The Northeast sees the sharpest rise, peaking near 28 deaths by 2016, while the South follows as the second-highest. The Midwest remains the lowest, and the West plateaus after an initial increase. This highlights regional disparities in the opioid epidemic's impact.
```{r plot 2, echo=FALSE}
# Install and load necessary libraries
suppressWarnings(suppressMessages({
  if (!require(plotly)) install.packages("plotly")
  library(plotly)
  if (!require(dplyr)) install.packages("dplyr")
  library(dplyr)
}))

# Filter and prepare the data
filtered_data <- drug_mortality %>%
  filter(`Race and Hispanic Origin` %in% c("All Races - All Origins", 
                                           "Non-Hispanic Black", 
                                           "Non-Hispanic White", 
                                           "Hispanic")) %>%
  group_by(Year, `Race and Hispanic Origin`) %>%
  summarize(mean_crude_rate = mean(`Crude Death Rate`, na.rm = TRUE), .groups = "drop")  # Average Crude Death Rate

# Create an interactive scatterplot 
deaths_by_race_ethnicity <- plot_ly(filtered_data, 
               x = ~Year, 
               y = ~mean_crude_rate, 
               type = 'scatter', 
               mode = 'lines+markers', 
               color = ~factor(`Race and Hispanic Origin`, levels = c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic")), # Reorder legend
               colors = "Set2",  # Use a color palette
               text = ~paste("Year:", Year, "<br>Rate:", mean_crude_rate),
               hoverinfo = "text") %>%
  layout(
    title = "Fig. 2: Association Between Race/Hispanic Origin and Drug Mortality (1999-2016)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Mean Crude Death Rate (per 100,000)"))


# Display the plot
deaths_by_race_ethnicity
```
**Figure 2 Description:** This line chart shows the association between race/ethnicity and drug-related mortality rates from 1999 to 2016, highlighting trends for Hispanic, Non-Hispanic Black, and Non-Hispanic White populations. Non-Hispanic Whites experienced the steepest increase in mortality, while Hispanics experienced the least steep increase.
```{r plot 3, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
suppressWarnings(suppressMessages({
  if (!require(plotly)) install.packages("plotly")
  library(plotly)
  if (!require(dplyr)) install.packages("dplyr")
  library(dplyr)}))

# Summarize the data to count deaths, stratified by Sex
deaths_by_sex <- drug_mortality %>%
  group_by(Year, Sex) %>%
  summarise(
    Total_Deaths = sum(Deaths, na.rm = TRUE), 
    Total_Population = sum(Population, na.rm = TRUE),
    .groups = "drop")

# Create the ggplot bar plot with deaths by sex and hover text
sex_vs_deaths_plot <- ggplot(deaths_by_sex, aes(
  x = Year, y = Total_Deaths, fill = Sex,
  text = paste(
    "Year:", Year,
    "<br>Sex:", Sex,
    "<br>Total Deaths:", Total_Deaths))) +
  geom_bar(stat = "identity") +
  labs(title = "Fig. 3: Total Count Drug Poisoning Deaths by Sex (1999-2016)",
       x = "Year", y = "Total Drug Poisoning Deaths") +
  facet_wrap(~ Sex) +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "skyblue", "Female" = "pink")) +
  scale_y_continuous(labels = scales::label_comma())  # Update y-axis to show whole numbers and add clarity for units

# Convert the ggplot to an interactive plotly plot and use the `text` aesthetic for tooltips
interactive_sex_vs_deaths_plot <- ggplotly(sex_vs_deaths_plot, tooltip = "text")

# Display the interactive plot
interactive_sex_vs_deaths_plot
```
**Figure 3 Description:** This bar chart illustrates the total number of drug poisoning deaths from 1999 to 2016, broken down by sex (Both Sexes, Female, Male). The data highlights a steady increase in deaths over time, with males consistently having higher counts.

```{r plot 4, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)
library(paletteer)
deaths_by_age <- drug_mortality %>%
  group_by(Year, `Age Group`) %>%
  summarise(
    Total_Deaths = sum(Deaths, na.rm = TRUE),
    Total_Population = sum(Population, na.rm = TRUE),
    .groups = "drop")

#Removal of All Ages value to see each age range's values
deaths_by_age_filtered <- deaths_by_age %>%
  filter(`Age Group` != "All Ages")

# Create the ggplot bar plot with deaths by Age Group and hover text
age_vs_deaths1_plot <- ggplot(deaths_by_age_filtered, aes(
  x = Year, y = Total_Deaths, fill = `Age Group`,
  text = paste(
    "Year:", Year,
    "<br>Age Group:", `Age Group`,
    "<br>Total Deaths:", Total_Deaths))) +
  geom_bar(stat = "identity") +
  labs(title = "Fig. 4: Total Count Drug Poisoning Deaths by Age Range (1999-2016)",
       x = "Year", y = "Total Drug Poisoning Deaths") +
  facet_wrap(~ `Age Group`) +
  theme_minimal() +
  scale_y_continuous(labels = scales::label_comma()) +  # Update y-axis to show whole numbers and add clarity for units
  scale_fill_paletteer_d("MetBrewer::Cassatt1")

# Convert to interactive plotly plot
interactive_age_vs_deaths1_plot <- ggplotly(age_vs_deaths1_plot, tooltip = "text")

# Display the interactive plot
interactive_age_vs_deaths1_plot
```
**Figure 4 Description:** This faceted bar chart displays the total drug poisoning deaths from 1999 to 2016, segmented by age groups. Middle age groups (25–34, 35–44, 45–54) show the steepest increases in total deaths, particularly 25–34 and 45–54, which display significant jumps post-2010. Those on the tail ends of this data (youngest and oldest age groups of the population) have the lowest rates, but still exhibit some drug poisoning death rates. 
